---
  output:
    github_document:
      df_print: kable
---
   
   <!-- README.md is generated from README.Rmd. Please edit that file -->
   
   ```{r knitr_init, echo=FALSE}
knitr::opts_chunk$set(
   collapse=TRUE,
   warning=FALSE,
   message=FALSE,
   comment="#>",
   fig.path="man/figures/README-"
);
```

Genejam is intended to freshen gene annotations to the
current official standard. It is particularly useful when
comparing genes in two datasets, for example when those datasets
may not be using the same gene symbols to represent equivalent
genes.

## Installation

Install using the `remotes` package:

`remotes::install_github("jmw86069/genejam")`

Note: It is recommended not to use the `devtools` package to
install Github packages, mostly because the `devtools` package
has many more components than are required for installation.
Instead, `devtools` includes all components needed to develop
R packages, beyond the scope of installing one such R package.

## freshenGenes()

The simplest use is to supply a set of gene symbols:

```{r, genejam_1}
genejam::freshenGenes(c("APOE", "APOA", "HIST1H1C"))
```

For slightly more detail, you can edit the argument `final`
to include things like `"SYMBOL"` (default), `"GENENAME"`,
`"ALIAS"`, `"ACCNUM"`, and more.

```{r, genejam_2}
genejam::freshenGenes(c("APOE", "APOA", "HIST1H1C"),
   final=c("SYMBOL", 
      "GENENAME",
      "ALIAS"))
```

## Convenience function freshenGenes2()

I frequently find myself wanting gene symbol, and the
long gene name, so I created a simple function `freshenGenes2()`
that uses default argument `final=c("SYMBOL", "GENENAME")`:

```{r, genejam_3}
genejam::freshenGenes2(c("APOE", "APOA", "HIST1H1C"))
```

## Convenience function freshenGenes3()

The other common use case is to include other gene aliases,
with the function `freshenGenes3()`:

```{r, genejam_4}
genejam::freshenGenes3(c("APOE", "APOA", "HIST1H1C"))
```


## Slightly more advanced

What if you already have Entrez gene ID, and want associated
annotations? The function `freshenGenes()` runs two steps:

1. convert input to `intermediate` (which is the Entrez gene ID)
2. convert `intermediate` to the output defined by `final`, for
example `final=c("SYMBOL")` would create a column `"SYMBOL"`.

In this example, the Entrez gene ID values are in a column
`"ENTREZID"`, so we will use argument `intermediate="ENTREZID"`.
In this case, you already have `"intermediate"`, so you invoke
the function with a `data.frame` with values in a column named
`"intermediate"`, and set `try_list=NULL`.

```{r, genejam_5}
df <- data.frame(ENTREZID=c("348", "4018", "3006", "100"));
genejam::freshenGenes2(x=df, intermediate="ENTREZID")
```

Similarly, you can provide input with a mixture of gene symbols
and Entrez gene ID values. Shown below is mixed input.

```{r, genejam_5a}
idf <- data.frame(Gene=c("MINA", "", "GABRR3", "GABRR3", ""),
   ENTREZID=c("", "84864", "", "200959", "200959"))
idf
```

You only need to specify `intermediate="ENTREZID"` as before.

```{r, genejam_5b}
genejam::freshenGenes2(x=idf, intermediate="ENTREZID")
```

Notice the values in `"ENTREZID"` are updated based upon the
first step resolution of `"Gene"` values to `"ENTREZID"`.
The `"SYMBOL"` and `"GENENAME"` columns are populated
using values in `"ENTREZID"`.


## When is genejam useful?

The official gene nomenclature is updated multiple times per year,
which means one Entrez gene ID may have a different official gene
symbol before and after the update. When comparing data
from two experiments, it is important to use the same
gene nomenclature. Otherwise, there will be differences in
results only because the names of some genes are different.

Most microarray platforms provide gene annotations,
which are updated much less frequently than the official
genes. For example, Affymetrix array "Clariom D Human"
was last updated between 2016 and 2018 (this document was
written in 2021.) In order to compare microarray results
to those from literature, biological pathways, or
other experiments, the gene nomenclature needs to
be updated to the most current version.

### Rationale for the workflow

In rare cases an official
gene symbol is "moved" from one Entrez gene ID to another,
usually when the original Entrez gene ID is deleted.
In these cases, the most reasonable link between
an experimental asay and the targeted gene is the
gene symbol. An alternative is to use a sequence
accession number used to design the assay.

As a result, a "best possible" gene annotation strategy
is used.

* When Entrez gene ID is available, use it.
* When Entrez official gene symbol is available, use that
to determine the Entrez gene ID.
* When a sequence accession used for assay design is available,
use that to determine the associated Entrez gene ID value
or values.
* When a gene symbol alias is available, use that
to determine the Entrez gene ID value or values
associated with this alias.

Sometimes an assay measures two genes. The steps in `genejam`
are designed to maintain multi-gene associations where
necessary. If one gene symbol alias is associated with
two (or more) genes, then all those genes will become
associated. Note this only happens if a higher priority
association was not already found.

Ultimately the workflow is what I and others have been doing
all along, to assemble the best available gene annotations
for a given dataset, while also leaving behind the fewest
possible un-annotated entries. When one source of annotation
fails, try another on missing entries; and so on.


## Optimizations

The steps used in `genejam` are designed for speed, to
the extent that providing 100,000 rows should return
results within a few seconds at most. Only unique
values are queried, and only missing entries are
updated. When multiple values are combined by a
delimiter, a highly optimized method is used.

Lastly, these operations also use optimized
mixed-alphanumeric sorting even in the context
of a `list`, so things like
`"chr2"` will appear before `"chr10"` in sort
order. Incidentally, a sort step is necessary
so you can compare whether two entries are
associated with the same genes. If one is stored
as `"APOE,APOE4"` and the other as `"APOE4,APOE"`,
this comparison fails.

All that to say, I use these functions a lot so
I need them to be reliable and fast. It takes a
few seconds just to load the associated SQLite
gene annotation data, and the process used by
`freshenGenes()` is usually substantially faster
than that step alone.


## Package Reference

A full online function reference is available via the pkgdown
documentation:
   
   Full genejam command reference:
   [https://jmw86069.github.io/genejam](https://jmw86069.github.io/genejam)
