---
   output: github_document
---
   
   <!-- README.md is generated from README.Rmd. Please edit that file -->
   
   ```{r knitr_init, echo=FALSE}
knitr::opts_chunk$set(
   collapse=TRUE,
   warning=FALSE,
   message=FALSE,
   comment="#>",
   fig.path="man/figures/README-"
);
```

Genejam is intended to freshen gene annotations to the
current official standard. It is particularly useful when
comparing genes in two datasets, for example when those datasets
may not be using the same gene symbols to represent equivalent
genes.

```{r, genejam_1}
genejam::freshenGenes(c("APOE", "APOA", "HIST1H1C"))
```

Specifically, in December 2019 the gene nomenclature for the large
histone gene family was changed, so the gene symbol `"HIST1H1C"` from
November 2019 is now called `"H1-2"` in December 2019 and thereafter.
These gene nomenclature updates happen frequently (perhaps monthly),
and apparently any gene symbol can be updated. So when we compare
a gene list created today, to any gene list not created today,
we run `freshenGenes()` on each gene list to ensure they both
use the same, "most current" gene annotation.


## Package Reference

A full online function reference is available via the pkgdown
documentation:
   
   Full genejam command reference:
   [https://jmw86069.github.io/genejam](https://jmw86069.github.io/genejam)
