---
  output:
    github_document:
      df_print: kable
---
   
   <!-- README.md is generated from README.Rmd. Please edit that file -->
   
   ```{r knitr_init, echo=FALSE}
knitr::opts_chunk$set(
   collapse=TRUE,
   warning=FALSE,
   message=FALSE,
   comment="#>",
   fig.path="man/figures/README-"
);
```

Genejam is intended to freshen gene annotations to the
current official standard. It is particularly useful when
comparing genes in two datasets, for example when those datasets
may not be using the same gene symbols to represent equivalent
genes.

## Installation

Install using the `remotes` package:

`remotes::install_github("jmw86069/genejam")`

Note: It is recommended not to use the `devtools` package to
install Github packages, mostly because the `devtools` package
has many more components than are required for installation.
Instead, `devtools` includes all components needed to develop
R packages, beyond the scope of installing one such R package.

## freshenGenes()

The simplest use is to supply a set of gene symbols:

```{r, genejam_1}
genejam::freshenGenes(c("APOE", "APOA", "HIST1H1C"))
```

For slightly more detail, you can edit the argument `final`
to include things like `"SYMBOL"` (default), `"GENENAME"`,
`"ALIAS"`, `"ACCNUM"`, and more.

```{r, genejam_2}
genejam::freshenGenes(c("APOE", "APOA", "HIST1H1C"),
   final=c("SYMBOL", 
      "GENENAME",
      "ALIAS"))
```

## Convenience function freshenGenes2()

I frequently find myself wanting gene symbol, and the
long gene name, so I created a simple function `freshenGenes2()`
that uses default argument `final=c("SYMBOL", "GENENAME")`:

```{r, genejam_3}
genejam::freshenGenes2(c("APOE", "APOA", "HIST1H1C"))
```

## Convenience function freshenGenes3()

The other common use case is to include other gene aliases,
with the function `freshenGenes3()`:

```{r, genejam_4}
genejam::freshenGenes3(c("APOE", "APOA", "HIST1H1C"))
```


## Slightly more advanced

What if you already have Entrez gene ID, and want associated
annotations? The function `freshenGenes()` runs two steps:

1. convert input to `"intermediate"` (which is the Entrez gene ID)
2. convert `"intermediate"` to the output defined by `final`, for
example `final=c("SYMBOL")`.

In this case, you already have `"intermediate"`, so you invoke
the function with a `data.frame` with values in a column named
`"intermediate"`, and set `try_list=NULL`:

```{r, genejam_5}
df <- data.frame(intermediate=c("348", "4018", "3006", "100"));
genejam::freshenGenes2(x=df, try_list=NULL)
```



## Comments on usefulness

Specifically, in December 2019 the gene nomenclature for the large
histone gene family was changed, so the gene symbol `"HIST1H1C"` from
November 2019 is now called `"H1-2"` in December 2019 and thereafter.
These gene nomenclature updates happen frequently (perhaps monthly),
and apparently any gene symbol can be updated. So when we compare
a gene list created today, to any gene list not created today,
we run `freshenGenes()` on each gene list to ensure they both
use the same, "most current" gene annotation.


## Package Reference

A full online function reference is available via the pkgdown
documentation:
   
   Full genejam command reference:
   [https://jmw86069.github.io/genejam](https://jmw86069.github.io/genejam)
